---
title: "Mapping benthic habitat sensitivity to fishing activities"
subtitle: ""
author: "Philip Haupt and Duncan Vaughan"
date: "29 March 2019"
output: html_document
bibliography: "references.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
The United Kingdom's (UK) benthic habitat is a critical component of the marine ecosystem which supports a rich fish and benthic invertebrate community. 

The benthic habitat is subject to a vast array of pressures in the marine environment, of which bottom towed fishing gears is the most prevalent and widespread activities.

Natural England provides statuary conservation advice on operations in the marine environment in National National Reserves (NNRs) in England, including terrestrial and marine environment. In the marine environment, NNRs include special Protected Areas (SPAs), Special Areas of Conservation (SACs), Marine Conservation Zones (MCZs), collectively referred to as Marine Protected Areas (MPAs). All MPAs in the England are currently multi-use areas, within which protected features exist. Protected features include habitats The call for wider seas conservation 



Fishing regulations may change in due process when the UK leaves the EU (EU exit) on the 31 March 2019, which is likely to have substantial ramifications for fisheries and fisheries management *(Appleby and Harrison, 2017)*. The Queen’s speech indicated that a new Fisheries Bill would be drafted to set domestic fishing quotas, and preserve fishing stocks (https://www.instituteforgovernment.org.uk/explainers/queens-speech). A change to fishing regulations requires that due consideration be given to the impacts of a displacement of fishing effort on the environment and to livelihoods of fishermen *(ABPmer, 2017; Guidetti, 2007)*. A displacement of fishing effort is defined here as the new or different areas where fishing effort moves to, which may be accompanied by a change in fishing gear used, in response to a change in fishing regulations (Vaughan, 2017). Previous research (studies) have shown that a displacement of fishing effort could lead to detrimental environmental and socio-economic effects (Bastardie et al., 2017; Chollet and Mumby, 2015).
The overarching aim of this project is provide advice to fisheries managers to minimise damage to UK marine ecosystems by avoiding negative impacts on the ecosystem arising from the displacement of fishing effort from the various [inshore] fisheries. Specifically, the project aims to identify recently non-fished sensitive benthic habitat, and essential fish habitat which are currently only lightly impacted, or not impacted. 





Habitat sensitivity mapping
Fishing displacement socio-economics [@Groot2014] [@Vaughan2017]
The EU Exit 
test citations: [@Bastardie2013; @Groot2014]

### Objectives
The objective of the project is to map the spatial distribution of sensitive benthic habitat in relation to 36 OSPAR standardized pressures caused by fishing activities in English territorial waters (12 NM offshore). 
 and their corresponding pressures to inform Natural England’s Conservation Advice. The study area includes the wider seas (outside of MPAs) to provide evidence of sensitive habitats which may be affected by a displacement of fishing pressures arising from different scenarios of EU Exit. 
A secondary aim for the project is develop a consolidated spatial index of the combined suite of pressures stemming from each fishing activity to provide a summarised overview of benthic habitat sensitivity to inform NE’s conservation advice. 
A third project objective is the development of maps of the sensitivity distribution for 30 fish and shellfish species within UK waters to fishing related activities. These models are underpinned by sensitivity assessments carried out at set benchmarks, and coupled with species distribution models.
A fourth objective is to link the work to policy on Sustainable Fishing within English territorial waters to provide high level direction to Natural England. Specifically this component will help develop the principles which inform advice on sustainable fishing in English territorial waters. This part of the project will attempt to develop a clear understanding of how sensitivity maps can be used to inform sustainable fishing policy and practice. 


Fishing displacement models have been developed previously [@Bastardie2013].

*TO DO: Section on biotopes and habitat, and the scale of information availbe.*

# Methods

## Study Area
Marine habitat sensitivity maps were generated for the United Kingdom's Exclusive Economic Zone (EEZ), from the shoreline to 200 NM offshore (Figure 1). The sensitivity maps were generated in two steps, first for 0 - 12 NM offshore for English territorial waters, and second for 12 - 200 NM offshore and 0 - 12 NM for devolved administrations. The sensitivity maps were generated independently for two geographical areas for which


```{r, echo = FALSE}
library(maps)       # Provides functions that let us plot the maps
library(mapdata)        # Contains the hi-resolution points that mark out the countries.

#map('worldHires',
#    c(),
#    xlim=c(-11,3), ylim=c(49,60.9))	

map('worldHires',
    c('UK', 'Isle of Man','Isle of Wight', 'Ireland'),
    xlim=c(-11,3), ylim=c(49,60.9),
    fill = TRUE,
    col = c("grey50", "grey80", "grey80", "grey80"))#,
    #add = TRUE)	
```
*TO DO replace with maps from QGIS?: Figure 1. Insert a map of the UK coastline and 200 NM Exclusive Economic Zone (EEZ) Or Map of 0 - 12NM and a second map of develoved admin*


### Data sources
Two main data sets were used to map the sensitivities for the marine benthic habitats within the UK's EEZ, include 1) habitat sensitivity assessments, and 2) a marine habitat map for the UK. The two data sources share common unique identifier of the habitat catogory which was assessed and that which appears on the habitat map, and could therefore be joined to generate habitat sensitivity maps.

#### 1. Marine biotope map
The standard benthic habitat map used by NE in its day to day operations, was used as the basis to inform the spatial distribution of benthic habitats around within the UK's EEZ for this project. The habitat cateogries contained within NE's habitat map is based on the European Nature Information System (EUNIS) classification developed in conjunction with the JNCC [@Davies2004; @Galparsoro2012]. 
*TO DO: Get detail from Ian/Tim on the habitat map - is there a NE document to reference?*

#### The EUNIS habitat classification system
The EUNIS benthic habitat categories is a hierarchical system within which fine-scale habitat categories are nested within broader scale habitat categories, referred to as EUNIS levels. At the top of the hierarchical classification system, the broadest definition of habitat corresponds, e.g. "Marine", to EUNIS level 1. At higher EUNIS values, the amount of thematic detail about the habitat increases progressively, for example, EUNIS level 2: A1 : Littoral rock and other hard substrata,
EUNIS level 3: A1.1 : High energy littoral rock,
EUNIS level 4: A1.11 : Mussel and/or barnacle communities,
EUNIS level 5: A1.113 : *Semibalanus balanoides* on exposed to moderately exposed or vertical sheltered eulittoral rock,
EUNIS level 6: A1.1132 : *Semibalanus balanoides*, *Fucus vesiculosus* and red seaweeds on exposed to moderately exposed eulittoral rock. 

See the [EUNIS-webpage](https://eunis.eea.europa.eu/habitats-code-browser.jsp) for more examples on the hierarchical view of the EUNIS habitat classification system.


The benthic habitat map was generated using a variety of methods and data sources, such as remote sensing, side-scan sonar, sediment grabs and SCUBA-diver surveys, through the collaboration of a large number of orgnaisations. As such, there is variability in the amount of detail about the habitat (substrate and benthos), as well as the confidence associated thereto for areas around the coastline. The EUNIS benthic habitat map is therefore comprised of a patchwork of polygons which differ in their shape, area and the  habitat category which they represent. The habitat categories are of different EUNIS levels, and the confidence associated with the patches of habitat are varied. The confidence associated with habitat clasifcation of each polygon is represented by a MESH score (*Annex MESH confidence*)

highest spatial resolution contained within the habitat map used by NE are mapped 

, will be used in conjunction with existing benthic habitat maps to derive sensitivity maps for benthic habitat.

#### 2. Sensitivity assessments
Sensitivity assessments of each betnhic habitat category has been carried out for all habitats.
Natural England's exiting conservation advice database, PD_AoO.accdb, houses sensitivity assessments for all biotopes in English territorial waters. The sensitivity assessments are based upon the MarESA [@Tillin2014]. methods. the methodology uses a standardised approach in which each biotope (formally recognised) was assessed against the *Pressures* arising from 



### Data processing
All technical work is being carried out using the R statistical environment in effort to make the work repeatable and provide instant results following updates to sensitivity assessments or updates to the benthic habitat map.

# Annex
**Table.** EUNIS MESH confidence scores

Confidence Range (%) | Confidence Level        | Explanation
---------------------|-------------------------|---------------------------------------------------
0 < x ≤ 20           | Very low confidence     | -
20 < x ≤ 37          | Low confidence          | Upper limit is the highest possible score for maps that use only physical ground truthing data
37 < x ≤ 58         |Moderate confidence     |Upper limit is the highest possible score for maps that use only physical and biological ground truthing data.
58 < x ≤ 79         |High confidence         |Upper limit is the highest possible score for maps that use only physical ground truthing and remote sensing data.
79 < x ≤ 100        |Very high confidence    |Maps with confidence scores higher than 79 % definitely include biological ground truthing and remote sensing data.
	
# References
