---
title: "Mapping benthic habitat sensitivity to fishing activities"
subtitle: "Natural England report XXX-draft"
author: "Philip Haupt and Duncan Vaughan"
date: "29 March 2019"
output: html_document
bibliography: "./references.bib"
fig_width: 6 
fig_height: 4 
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(maps)       # Provides functions that let us plot the maps
library(mapdata)        # Contains the hi-resolution points that mark out the countries.
library(png)
```   

# Introduction

The United Kingdom's (UK) marine environment is of significant economic value, in which the exploitative activities and infrastructure development is constantly weighed up against their impact on the environment [@walportForesightFutureSea2017].  The UK's marine environment is subject to a vast array of pressures, such as fishing operations, marine aggregate dredging, development of renewable energy and risks of pollution from oil spills among many others. Overfishing and habitat loss are regarded as two of the main contributors to the degradation of the UK's marine environment [@walportForesightFutureSea2017].

Fishing operations are widespread throughout the United Kingdom, and is fished by both national and international fleets within the remits of the The London Fisheries Convention (1964) and the Common Fishery Policy (CFP 1970s). Many of the existing regiulations and laws are therefore enshrined in EU maritime law. The withdrawal of UK from the European Union (Brexit), which is scheduled for the 29 March 2019,[European Union (Withdrawal) Act 2018, section 20](http://www.legislation.gov.uk/ukpga/2018/16/section/20); [Article 50](https://eur-lex.europa.eu/legal-content/EN/TXT/HTML/?uri=CELEX:12016M/TXT&from=EN), is likely to lead to change in fisheries regulations and management [@phillipsonSeaTroublesBrexit2018]. Brexit may influence how fish stocks are shared across political borders, quota allocation to EU-member states in the UK, and Total Allowable Catch (TAC) for different stocks among many other aspects related to managing the UK fishing grounds [@phillipsonSeaTroublesBrexit2018]. A change in fishing regulations and management is therefore likely to have substantial ramifications fishing operations in the UK. The spatial distribution of fishing effort among various fishing metiers is a key element is likely to be influenced by a change in regulations [@vaughanFishingEffortDisplacement2017]. Previous research has shown that a displacement of fishing effort could lead to detrimental environmental and socio-economic effects [e.g. @stevensonSocioeconomicConsequencesFishing2013a]. It is therefore important that the impact on the marine environment brought about by a displacement of fishing effort be considered prior to adopting new regulations or change in management practices [@vaughanFishingEffortDisplacement2017; @chollettQuantifyingSqueezingStretching2016]. 

Delineating the geographical distribution of habitat sensitivity has been identified as critical dataset for managing human activities on the marine environment [Bax, N.J. & Williams, A. ( 2001) Seabed habitat on the south‐eastern Australian continental shelf: context, vulnerability and monitoring. Marine and Freshwater Research, 52, 491– 512.](http://0-www.publish.csiro.au.wam.seals.ac.za/mf/MF00003). The sensitivity of marine habitats have previously been defined as sensitivity as *"the degree to which marine features respond to stresses, which are deviations of environmental conditions beyond the expected range"* [Sensitivity and Vulnerability in Marine Environments: an Approach to Identifying Vulnerable Marine Areas](https://0-onlinelibrary.wiley.com.wam.seals.ac.za/doi/full/10.1111/j.1523-1739.2005.00148.x). Previous attempts to use quantitative methods to delineate benthic habitat sensitivity  were limited by the level of detail about the underlying benthic habitat upon which fishing activities impact. These studies have assumed that benthic habitat were comprised of four broad-scale categories, such as mud, muddy sand, sand and gravel. Current quantitiative models therefore do not consider differences in living structures occuring within these habitats. These models are therefore likely to over- or under-estimate the sensitivity of fine-scale habitats within these very broad defintitions of habitat. Previous alternative efforts to map out geographical distribution of habitat sensitivity have been critisised for being score based or expert judged systems e.g. Tyler Walters 1999, Zacharias and Greg 2005, which are therefore not repeatable or quantitative [Assessing and predicting the relative ecological impacts of disturbance on habitats with different sensitivities](https://0-besjournals.onlinelibrary.wiley.com.wam.seals.ac.za/doi/full/10.1111/j.1365-2664.2007.01274.x); [A size-based model of the impacts of bottom trawling on benthic community structure](https://www.nrcresearchpress.com/doi/10.1139/f02-148#.XIpjkSj7RPY). Despite this limitation, both quantitative and scored based systems of habitat sensitivity have been based upon the same central tenet of resilience and recovery.

Natural England provides statuary conservation advice on operations in English territorial waters (from the shore to 12 NM offshore) to the UK Government and industry (and out to 200 NM offshore for renewable energy). Natural England uses sensitivity assessments of protected bethic habitats and species to inform its advice on operations. The sensitivity assessments include 39 pressures associated with a 100 activities stemming from 18 operations. Natural England's statuatory remit is restricted to protected habitats and species within English marine protected areas (MPAs), known as designated features. The habitat component of the designated features are defined at a intermediate scale of detail. The conservation advice is therefore limited to a subset of the area of English territorial waters, and provided at a coarser scale of habitat detail than is available for the UK seabed. In recognition that impacts from a displacement of fishing effort occurring outside of MPAs can have an impact on features within MPAs, a "wider-seas" and "whole-sites" approach was used to assess the sensitivity of benthic habitats to fishing activities for the UK's EEZ (Exclusive Economic Zone). The whole-sites and wider-seas approach means that currently protected features were not considered in isolation, but that the connectivity between benthic habitats were considered to provide a holistic overview the spatial distributions of fishing related impacts on the benthic habitat. 

The objective of this project was to map the spatial distribution of sensitive benthic habitats in relation to 39 pressures caused by fishing activities in the UK's Exclusive Economic Zone (EEZ). The study area includes the wider seas (outside of MPAs) to provide evidence of sensitive habitats which may be affected by a displacement of fishing pressures arising from different scenarios of EU Exit. 

Previous studies have developed tools to predict the impacts of fishing effort displacement, which forecast a change in fishing effort displacement [@bastardieDISPLACEDynamicIndividualbased2013]. This project did not attempt to predict a change in fishing activity given the significant uncertainty and complexity in how Brexit would affect fishing opportunities and regulations in the UK. In this study habitat sensitivity to standardised pressures from various fishing activity were mapped as a precautionary approach given the manifold uncertainties associated with fishing displacement arising from Brexit, such as how trade negotiations may influence fishing rights allocations. Previous work which have mapped benthic habitat sensitivity have used a different approach . Broad-scale benthic habitat and

A case study was developed to explore the potential use of benthic habitat sensitivity models to help inform NE of potential impacts ion the marine environment. 

In this report a case study was developed to inspect the potential application of the habitat sensitivity mapping in relation to spatial maps of fishing activity. Fishing activities have had profound effects on the UK's marine ecosystem, of which bottom towed fishing gears is one of the most damaging activities which altered benthic habitats, composition and trophic structure of fish and invertebrate communities over the last century of industrial fishing [@kaiserChangesMegafaunalBenthic1998; @jenningsImpactsTrawlingDisturbance2001; @hiddinkCumulativeImpactsSeabed2006; @tillinChronicBottomTrawling2006; @thurstanEffects118Years2010]. The spatial overlap of benthic habitat sensitivity maps and the distribution of fishing related dredging activities were compared to evaluate the use of sensitivity mapping as a tool to inform management of options to minimise the impacts of a the spatial distribution of fishing activities. The habitat sensitivity maps were therefore evaluated as an data set to inform conservation advice in the event of displacement of fishing effort. The ability of the tool to identify recently non-fished sensitive benthic habitats were explored, as a means to inform spatial planning in the UK's EEZ. 

#### Remove: 
The equitable redistribution of marine resources, such as fishing quota allocation, remains a key challenge to resource managers, for example. [@halpernConfoundingEffectsExport2004].
The remit of NE's statutory powers : National National Reserves (NNRs) in England, including terrestrial and marine environment. In the marine environment, NNRs include special Protected Areas (SPAs), Special Areas of Conservation (SACs), Marine Conservation Zones (MCZs), collectively referred to as Marine Protected Areas (MPAs) hereafter. All MPAs in the England are currently multi-use areas, within which protected features exist.
Remove: It is therefore critical for marine resource managers to understand the distribution of fishing impacts in relation to the distribution the habitats and species to minimise damage to UK marine ecosystems. It is therefore imperative that marine environment and fisheries bodies make the necessary provisions to minimise potential damage arising from a change in fishing regulation following Brexit.

# Methods

## Study Area
Marine habitat sensitivity maps were generated for the United Kingdom's Exclusive Economic Zone (EEZ), from the shoreline to 200 NM offshore [Figure 1]!(english_waters). The sensitivity maps were generated in two steps, first for 0 - 12 NM offshore for English territorial waters, and second, comprising both the 12 - 200 NM English offshore and 0 - 200 NM for devolved administrations. The habitat sensitivity maps were therefore generated independently for these adjacent geographical areas. 

The areas were treated independently as Natural England holds fine-scale data on the composition of the benthic habitat in 0 - 12 NM in England which was not available for the offshore (12 - 200 NM) English waters and 0 - 200 NM for Devolved Administrations at the time of writing. The fine-scale information on habitat composition was utilised in an additional step within the R modelling routines, which allowed biotope prediction within mapped broader scale habitats. The biotope level habitat information was obtained from additional data and expert validation which describes the likely biotopes within subbiogeographic regions within English waters *insert a map of subbiogeogrpahic regions*. (This fine-scale habitat information when NE provides statutory advice, and is therefore well established within the organisations day to day working.)


```{r english_waters, echo = FALSE}
#map('worldHires',
#    c(),
#    xlim=c(-11,3), ylim=c(49,60.9))	

map('worldHires',
    c('UK', 'Isle of Man','Isle of Wight', 'Ireland'),
    xlim=c(-11,3), ylim=c(49,60.9),
    fill = TRUE,
    col = c("grey50", "grey80", "grey80", "grey80"))#,
    #add = TRUE)	
```
*TO DO replace with maps from QGIS? or add the eez?: Figure 1. Insert a map of the UK coastline and 200 NM Exclusive Economic Zone (EEZ) Or Map of 0 - 12NM and a second map of develoved admin*

### Data sources
Two main data sets were used to map the sensitivities for the marine benthic habitats within the UK's EEZ, including 1) habitat sensitivity assessments, and 2) a marine habitat map for the UK. The two data sources share a common unique identifier which refers to the habitat category and could therefore be joined to associate the sensitivity of habitat categories to the mapped habitat categories.

#### 1. Marine habitat map
The benthic habitat map used by NE (Marine_Evidence_Base_Internal.gdb, layer = BSH), was used as the basis to inform the spatial distribution of benthic habitats around within the UK's EEZ for this project. The habitat categories contained within NE's habitat map are based on the European Nature Information System (EUNIS) classification developed in conjunction with the JNCC [@Davies2004; @Galparsoro2012]. 
*TO DO: Get detail from Ian/Tim on the habitat map - is there a NE document to reference?*

##### The EUNIS habitat classification system
The EUNIS benthic habitat categories is a hierarchical system within which fine-scale habitat categories are nested within broader scale habitat categories, referred to as EUNIS levels. At the top of the hierarchical classification system, the broadest definition of habitat corresponds, e.g. "Marine", to EUNIS level 1. At higher EUNIS values, the amount of thematic detail about the habitat increases progressively, for example, EUNIS level 2: A1 : Littoral rock and other hard substrata,
EUNIS level 3: A1.1 : High energy littoral rock,
EUNIS level 4: A1.11 : Mussel and/or barnacle communities,
EUNIS level 5: A1.113 : *Semibalanus balanoides* on exposed to moderately exposed or vertical sheltered eulittoral rock,
EUNIS level 6: A1.1132 : *Semibalanus balanoides*, *Fucus vesiculosus* and red seaweeds on exposed to moderately exposed eulittoral rock. 

*See the* [EUNIS-webpage](https://eunis.eea.europa.eu/habitats-code-browser.jsp) *for more examples on the hierarchical view of the EUNIS habitat classification system.*

The habitat map was comprised habitat categories which were of different EUNIS levels (2 - 6). The differences in the habitat categories (EUNIS levels) is a result of different amounts of data available for the different areas across the EEZ. The development of the habitat is an ongoing process which updates the habitat maps and habitat classification system as new information becomes available. As such, there is variability in the amount of detail about the habitat (substrate and benthos), as well as variability in  confidence between the mapped areas (referred to as polygons hereafter). The habitat map is therefore comprised of a patchwork of polygons which may differ in their shape, area and the habitat category (and EUNIS level) which they represent, as well as the level of confidence with which the habitat was category was assigned to it. The confidence associated with habitat classification of each polygon is represented by a MESH (Mapping European Seabed Habitats) score (*Annex MESH confidence*). MESH scores are derived from the confidence associated with different data sources used to inform the classification of the habitat category assigned to a polygon. The data sources used to generate the benthic habitat map are based on a variety of methods and data sources, such as remote sensing, side-scan sonar, sediment grabs and SCUBA-diver surveys.

The EUNIS benthic habitat categories which appear on the habitat map were associated with habitat categories contained within the sensitivity assessment database, which were used together to derive sensitivity maps for benthic habitat.

#### 2. Sensitivity assessments
Sensitivity assessments of each benthic habitat categories has been carried out for most EUNIS habitats, and are publicly available on The Marine Life Information Network's (MarLIN) [website](https://www.marlin.ac.uk/).The sensitivity assessments are based upon the MarESA [@tillinAssessingSensitivitySubtidal2014] methods in which the EUNIS habitats were assessed against 39 standardised pressures for each activity (Annex Activities and Pressures). The ten fishing activities in the AoO database were included in this project (Table 1). The pressures are standardised according to benchmarks, for which the benchmark value is defined at which point an effect is expected [@tillinAssessingSensitivitySubtidal2014]. The 39 pressures include the 36 OSPAR standardised pressures and an additional three which allow specifying direction of a pressure rather than simply stating a change (Table 2). For example instead of one pressure stating a change of 0.5&deg C, two separate pressures are included stating an increase of 0.5&deg; C and another a decrease of 0.5&deg; C. For details about the benchmarks please see Tillin et al  [-@tillinAssessingSensitivitySubtidal2014]. One of six levels of sensitivity are assigned to each habitat category in relation to the each of the 39 pressures (Table 3).


```{r fishing_acts, fig.cap="Table 1. Fishing activities", echo = FALSE}
library(RODBC)
library(DBI)
library(tidyverse)

#power pc path to database
db.path <- "D:/projects/fishing_displacement/2_subprojects_and_data/5_internal_data_copies/database/PD_AoO.accdb"
drv.path <- "Microsoft Access Driver (*.mdb, *.accdb)" #"this relies on the driver specified above for isntallation, and will not work without it!

#Below prints the list of options for the user to read, and then make a selection to enter below
#see key below
rpt.dir <- getwd()
setwd("F:/projects/marine_biotope_sensitivity")


db.p  <-  db.path
drv.p <-  drv.path

#specificy the connection to the database
connection.path <- paste0("Driver={",drv.path,"};DBQ=",db.path)#server=",srv.host,"; #server may need to be defined for other SQL database formats
        
# Connect to Access db to allow reading the data into R environment.
conn <- odbcDriverConnect(connection.path)
        
# Get individual data tables
        #----------------------------------------------------------------------------------------
#sql statements to translate Microsoft Access table into R objects - these will be needed make joined tables to arrive at the required datatable
tblActivityLUT <- sqlQuery(conn, paste("SELECT tblActivityLUT.*
                                       FROM tblActivityLUT;"))
tblActivityLUT <- as.data.frame(sapply(X = tblActivityLUT, FUN = as.character), stringsAsFactors=FALSE)
tblActivityLUT$OperationCode <- as.integer(tblActivityLUT$OperationCode)
        
tblActivityLUT %>% 
        filter(OperationCode == 11) %>%
        select(ActivityName, Description) %>%
        kable(caption = "Table 1: Fishing activities in Advice on Operations (AoO) database") %>%
        kable_styling(position = "left", full_width = F)

close(conn)
setwd(rpt.dir)
```

``` {r pressures, fig.cap="Table 2. The 39 standrdised pressures against which habitat sensitivities were assessed", echo = FALSE}
library(RODBC)
library(DBI)
library(tidyverse)

#power pc path to database
db.path <- "D:/projects/fishing_displacement/2_subprojects_and_data/5_internal_data_copies/database/PD_AoO.accdb"
drv.path <- "Microsoft Access Driver (*.mdb, *.accdb)" #"this relies on the driver specified above for isntallation, and will not work without it!

#Below prints the list of options for the user to read, and then make a selection to enter below
#see key below
rpt.dir <- getwd()
setwd("F:/projects/marine_biotope_sensitivity")


db.p  <-  db.path
drv.p <-  drv.path

#specificy the connection to the database
connection.path <- paste0("Driver={",drv.path,"};DBQ=",db.path)#server=",srv.host,"; #server may need to be defined for other SQL database formats
        
# Connect to Access db to allow reading the data into R environment.
conn <- odbcDriverConnect(connection.path)
        
# Get individual data tables
        #----------------------------------------------------------------------------------------
#sql statements to translate Microsoft Access table into R objects - these will be needed make joined tables to arrive at the required datatable
tblPressureLUT <- sqlQuery(conn, paste("SELECT tblPressureLUT.*
                                                FROM tblPressureLUT;"))
tblPressureLUT <- as.data.frame(sapply(X = tblPressureLUT, FUN = as.character), stringsAsFactors=FALSE)

tblPressureLUT %>% 
        kable(caption = "Table 2: Standardised 39 ppressures in Operations (AoO) database") %>%
        kable_styling(position = "left", full_width = F)

close(conn)
setwd(rpt.dir)
```

```{r sens_levels, fig.cap="Table 3. The six sensitivity levels which were assigned to habitats in relation to the 39 standardised pressures.", echo = FALSE}
#library(tidyverse)
sens.rank <- tribble(
        ~Rank, ~Sensivity_description,
        #-------/------------
        1, "Not sensitive",
        2, "Insufficient evidence",
        3, "Not assessed",
        4, "Low",            
        5, "Medium",
        6, "High"
)

sens.rank %>%
        kable(caption = "Table 3. The six sensitivity levels which were assigned to habitats in relation to the 39 standardised pressures.") %>%
        kable_styling(position = "left", full_width = F)

```
##### 2.1 Sensitvity assessments database
The sensitivity data is also stored internally at Natural England in a Microsoft Access database, namely, PD_AoO.accdb. the sensitivity assessments are used to inform NE's Advice on Operations (AoO), available as conservation advice packages to the public on the [Designated Sites View](https://designatedsites.naturalengland.org.uk/Marine/MarineSiteDetail.aspx?SiteCode=UKMCZ0045&SiteName=aln&countyCode=&responsiblePerson=&SeaArea=&IFCAArea=). The Microsoft Access database, PD_AoO.accdb, was used to obtain sensitivity assessments for the EUNIS habitats in this project. 

# START HERE

### Expert fine-scale benthic habitats within inshore sub-biogeoregions 
Fine-scale benthic habs

### 3. Data processing
A few GIs steps were required, and all other technical data manipulation was carried out using the R statistical environment in effort to make the work repeatable and provide instant results following updates to the sensitivity assessment database (PD_AoO.accdb) or updates to the benthic habitat map.



#### 3.1 GIS steps


step 1
rebuild the sbgr layer using set SNAP tolerance to 20m to inshore boundary of existing sbgr layer, and then extend outer edges beyond current extent. sbgr (Keith Hiscock) layer

Step 2

clip (difference) step 2 by the 12 NM limit mask (Neil Irvine), and the layer from step 1.

Step 3. 
Join (one to one) the properties of the sbgr (kH) layer to the newly clipped layer to obtain the correct attributes, and carried out a few checks to see that the correct associations were made. 
Output: D:\projects\fishing_displacement\2_subprojects_and_data\2_GIS_DATA\biogeoregions\subbiogeoregions_clipped12nm_joined_sbgr_PH_wgs84.shp


GIS habitat map steps

Production of the Offshore and Inshore areas.

Offshore (outside 12 NM) habitat map
F:\copy_data\Marine_Evidence_Base_Internal.gdb → CLIP with D:\projects\fishing_displacement\2_subprojects_and_data\2_GIS_DATA\masks\mask_minus_12nm_and_land.shp

Inshore (inside 12NM) habitat map
F:\copy_data\Marine_Evidence_Base_Internal.gdb → DIFFERENCE with D:/projects/fishing_displacement/2_subprojects_and_data/2_GIS_DATA/masks/mask_minus_12nm_and_land_WGS84.shp
NAME: D:\projects\fishing_displacement\2_subprojects_and_data\2_GIS_DATA\Marine habitat\Input_Polys_WGS84_Internal_BGR_inside_12nm.gpkg


Associate subbiogeoregions to Inshore habitat map
Input_Polys_WGS84_Internal_BGR_inside_12nm.gpkg → INTERSECT with  subbiogeoregions_clipped12nm_joined_sbgr_PH_wgs84.shp

#### 3.2 R Main code steps

githib link







# Case study materials methods
The habitat sensitivity map was developed for abrasion from dredging for the UK's EEZ, using the two step process described above. 

GIS operations
A fishing pressure layer for dredging and the current MPA layer was overlaid onto the habitat sensitivity map to investigate 1.) the overlap of dredging activities on habitats which are sensitive thereto, 2. ) the placement of MPAs in relation to the impacts on sensitive benthic habitats.

Fishing pressure layer

Marine protected area layer

# Results

In total 390 benthic habitat sensitivity maps were produced for 39 pressures from the ten fishing activities.
Output maps were vector based, and mirror the spatial resolution and habitat detail of the benthic habitat map.

## Case study results

1. The spatial distribution of benthic habitat sensitivity to abrasion from dredging is shown in Figure \@ref(fig:insh_drdg). (or merge the inshore and offshore into 1 figure perhaps?)

```{r insh_drdg, fig.cap= "Figure 2: The habitat sensitivity to abrasion from inshore dredging.",  echo = FALSE}
insh_fishing_dredging_img_path <- "F:/projects/marine_biotope_sensitivity/outputs/Fishing_dredging__Abrasion_disturbance.png"
#insh_fishing_dredging_img <- readPNG(insh_fishing_dredging_img_path, native = TRUE, info = TRUE)
#attr(fishing_dredging_img, "info")
#knitr::opts_chunk$set(fig.width=12, fig.height=8) 
knitr::include_graphics(insh_fishing_dredging_img_path, dpi = 150)
```
2. Overlay fishing activity
And provide some analysis of the proportion of sensitivity categories which overlap with dredging activity (which may need to be categorised into high/medium/low?) 
3. Overlay the MPAs, and repeat the above?


# Discussion

The sensitivity maps 

## Future development of the research
In a second phase essential fish habitat which are currently only lightly impacted, or not impacted will be identified.

A follow up project will consider: 
A third project objective is the development of maps of the sensitivity distribution for 30 fish and shellfish species within UK waters to fishing related activities. These models are underpinned by sensitivity assessments carried out at set benchmarks, and coupled with species distribution models.


# Annex
**Table.** EUNIS MESH confidence scores

Confidence Range (%) | Confidence Level        | Explanation
---------------------|-------------------------|---------------------------------------------------
0 < x ≤ 20           | Very low confidence     | -
20 < x ≤ 37          | Low confidence          | Upper limit is the highest possible score for maps that use only physical ground truthing data
37 < x ≤ 58         |Moderate confidence     |Upper limit is the highest possible score for maps that use only physical and biological ground truthing data.
58 < x ≤ 79         |High confidence         |Upper limit is the highest possible score for maps that use only physical ground truthing and remote sensing data.
79 < x ≤ 100        |Very high confidence    |Maps with confidence scores higher than 79 % definitely include biological ground truthing and remote sensing data.
	
#Refs to add
Tyler‐Walters, H., and Jackson, A. 1999. Assessing seabed species and ecosystems sensitivities; rationale and user guide. Report to English Nature, Scottish Natural Heritage and the Department of the Environment Transport, and the Regions from the Marine Life Information Network (MarLIN report 4). Marine Biological Association of the United Kingdom , Plymouth .

# References
